% §1 Introduction
% Status: Revised 2026-03-20 — full draft with empirical results from E1–E7.

\section{Introduction}
\label{sec:introduction}

Modern exascale systems expose unprecedented hardware heterogeneity---NVIDIA,
AMD, and Intel GPUs with distinct memory hierarchies, execution models, and
programming interfaces. Programming these systems productively requires
abstraction; programming them efficiently often demands its abandonment. Despite
a decade of investment in portability frameworks, application developers at the
Exascale Computing Project still rely on native or custom implementations in
approximately 55\% of production codes~\cite{Evans2022}, and performance
portability coefficients for leading frameworks collapse to values as low as
0.20--0.26 on certain architectures~\cite{Godoy2023}. The gap between the
promise of write-once, run-anywhere portability and the reality of
production-grade HPC software remains wide, consequential, and poorly
understood.

The community has responded with a rich ecosystem of solutions: portability
layers such as Kokkos, RAJA, and SYCL that abstract hardware
targets~\cite{Evans2022,heroux2020ecp}, task-based runtimes such as
Legion, PaRSEC, and Charm++ that shift scheduling and synchronization
responsibility into the runtime system~\cite{bauer2012legion,john2023multgpu,chandrasekar2023adaptive},
and translation-based approaches that preserve existing
toolchains~\cite{zhao2023hiplz}. Each approach reduces some dimension of
programming complexity. High-level language approaches such as Julia further
demonstrate that productivity and performance need not be mutually
exclusive~\cite{Godoy2023}. Yet none of these approaches
provides a principled answer to the question every developer faces in practice:
for \emph{this workload}, on \emph{this architecture}, at \emph{this
performance target}---should I use an abstraction or not?

What is missing is not another abstraction layer---it is the measurement
infrastructure to evaluate existing ones. Abstraction selection today is
heuristic, experience-driven, and opaque: developers rely on trial-and-error,
folklore, and vendor documentation rather than evidence. No systematic,
cross-layer empirical methodology exists for characterizing when abstraction
helps, when it fails, and why---across workload types, hardware architectures,
and backend maturity levels. This absence is not merely an inconvenience; it is
a structural barrier to informed software engineering at exascale. As the
community prepares for post-exascale systems with even greater architectural
diversity, the cost of this gap will only increase. The stakes are concrete: in
our measurements, Julia/\textsc{cuda.jl} delivers $1.25\times$ native DGEMM
throughput via a memory layout advantage (P005), yet collapses to $0.35\times$
efficiency on narrow BFS frontiers through per-launch dispatch amplification
(P008)---the same framework producing opposite outcomes on adjacent workloads,
depending solely on problem structure. On AMD CDNA3 (MI300X), Julia's
column-major layout advantage disappears entirely: the same Julia/ROCm kernel
delivers only $0.31\times$ native DGEMM throughput (P005 reversal), confirming
that abstraction outcomes are architecture-specific, not merely workload-specific.

This paper addresses this gap by presenting a \emph{measurement-driven
framework for selective abstraction} in GPU-accelerated HPC software. The
framework systematically characterizes abstraction effectiveness across a
three-dimensional experimental matrix: seven representative workloads spanning
regular (STREAM Triad, DGEMM, 3D Stencil), semi-irregular (SpMV, SpTRSV), and
irregular (BFS, N-body) computational patterns; five abstraction layers (Kokkos,
RAJA, SYCL, Julia, Numba) evaluated against native CUDA/HIP/SYCL baselines; and three
GPU architectures covering NVIDIA Blackwell (RTX 5060), AMD CDNA3 (MI300X), and
Intel Ponte Vecchio (GPU Max 1550). Using a four-level measurement hierarchy---performance
portability coefficients~\cite{Pennycook2019}, overhead attribution, root cause
analysis, and productivity metrics---we construct an empirical taxonomy of
abstraction failure modes and success patterns, and derive a quantified decision
framework that transforms abstraction selection from an art into an engineering
discipline. This work does not propose a new programming model or runtime
system; its contribution is the methodology and evidence base the community
needs to make informed decisions about existing ones.

Specifically, this paper answers three research questions:
\begin{itemize}
    \item[\textbf{RQ1}] \textit{When does abstraction help?} Under what workload
    characteristics and architectural conditions do high-level programming
    abstractions deliver on their portability promises without significant
    performance penalties?

    \item[\textbf{RQ2}] \textit{When does abstraction fail?} What are the
    systematic patterns of abstraction breakdown, and which layers---compiler,
    runtime, memory model---contribute to performance loss?

    \item[\textbf{RQ3}] \textit{How can we quantify the trade-off?} Can
    measurable, empirically-derived thresholds replace heuristic judgment in
    abstraction selection, enabling a reproducible and transferable decision
    framework?
\end{itemize}

Our main contributions are:

\begin{enumerate}
    \item An \textbf{empirical taxonomy} --- the first systematic classification
          of abstraction failure modes and success patterns across architectures
          and workloads (Section~\ref{sec:taxonomy}).
    \item A \textbf{measurement methodology} --- a reproducible protocol for
          cross-layer performance attribution
          (Section~\ref{sec:methodology}).
    \item A \textbf{quantified decision framework} --- evidence-based thresholds
          and rules for abstraction selection
          (Section~\ref{sec:framework}).
    \item A \textbf{performance portability dataset} --- 315 planned
          configurations across 5 abstraction layers, 3 GPU architectures,
          and 7 workloads, with full reproducibility artifacts.
    \item A \textbf{decision support tool} (\texttt{abstraction-advisor}) --- a
          command-line advisor that ingests workload profiles and recommends
          abstraction strategies.
\end{enumerate}

The remainder of this paper is organized as follows.
Section~\ref{sec:background} provides background on performance portability
metrics and the concept of selective abstraction.
Section~\ref{sec:related} reviews related work across exascale programming
practice, task-based runtimes, adaptive systems, and portability layers.
Section~\ref{sec:methodology} presents the measurement methodology and
experimental design.
Sections~\ref{sec:results} and~\ref{sec:taxonomy} present experimental results
and the resulting empirical taxonomy of failure and success patterns.
Section~\ref{sec:framework} presents the quantified decision framework.
Section~\ref{sec:validation} validates the framework against production
exascale application codes.
Section~\ref{sec:conclusion} concludes and outlines future directions.
