% §7 Decision Framework
% Status: Written 2026-03-26 — full rewrite with calibrated R1–R5 rules.

\section{Decision Framework}
\label{sec:framework}

The taxonomy of Section~\ref{sec:taxonomy} identifies eight abstraction
patterns (P001--P008) and their structural triggers. This section
operationalizes those patterns into a quantified decision framework: five
calibrated rules (R1--R5) that predict abstraction PPC from measurable
workload characteristics \emph{before} an abstraction is adopted. The
framework replaces heuristic judgment with evidence-based prediction,
answering RQ3: \textit{Can measurable, empirically-derived thresholds replace
heuristic judgment in abstraction selection?}

% ------------------------------------------------------------------
\subsection{Predictor Variables}
\label{subsec:predictors}
% ------------------------------------------------------------------

The framework requires five workload characteristics as input
(Table~\ref{tab:predictors}). All five are measurable from a native baseline
implementation and the workload's structural properties --- none require
running the abstraction first, which is what makes the framework predictive
rather than retrospective.

\begin{table}[t]
\centering
\caption{Predictor variables for the decision framework.}
\label{tab:predictors}
\begin{tabular}{llll}
\hline
\textbf{Variable} & \textbf{Symbol} & \textbf{Unit}
& \textbf{How to Measure} \\
\hline
Kernel execution time & $T_k$ & \textmu{}s
& GPU event timing on native kernel; median of 30 runs \\
Dependency levels & $L$ & ---
& CPU-side BFS on matrix/graph level structure \\
Arithmetic intensity & $AI$ & FLOP/byte
& \texttt{ncu --metrics} or empirical FLOP/memory ratio \\
Load imbalance index & $\sigma$ & ---
& $\text{std}(\text{row\_lengths}) / \text{mean}(\text{row\_lengths})$ \\
API completeness & $\text{API}_{\text{ok}}$ & boolean
& Can the abstraction express shared-memory tiling? \\
\hline
\end{tabular}
\end{table}

The API completeness variable deserves elaboration: it is a static property
of the abstraction, not the workload. Table~\ref{tab:api_ok} lists the
classification for each abstraction evaluated in this study.

\begin{table}[t]
\centering
\caption{API completeness by abstraction. \textit{True} indicates the
abstraction provides portable shared-memory allocation primitives sufficient
for tiled compute kernels (e.g., DGEMM).}
\label{tab:api_ok}
\begin{tabular}{lll}
\hline
\textbf{Abstraction} & \textbf{API$_{\text{ok}}$} & \textbf{Mechanism} \\
\hline
Kokkos (\texttt{TeamPolicy})     & True  & Scratch-pad memory via \texttt{ScratchMemorySpace} \\
SYCL (\texttt{local\_accessor})  & True  & \texttt{local\_accessor<T>} for workgroup-local memory \\
RAJA (\texttt{forall})           & False & No portable shared-memory allocation API \\
Julia (\texttt{@cuda}/\texttt{@roc}) & False & No scoped shared-memory primitives \\
\hline
\end{tabular}
\end{table}

% ------------------------------------------------------------------
\subsection{Decision Rules}
\label{subsec:rules}
% ------------------------------------------------------------------

The framework comprises five rules, presented here in their evaluation
priority order (highest priority first). Each rule is triggered by a
measurable condition, produces a PPC prediction from a closed-form formula,
and recommends an action.

% --- R3 ---
\paragraph{R3: API Expressivity Gate (P004).}
\textit{Priority: Highest.}

\noindent\textbf{Condition.}
$\text{requires\_shared\_memory\_tiling} = \text{true}$ \textbf{and}
$\text{API}_{\text{ok}}[\text{abstraction}] = \text{false}$.

\noindent\textbf{Predicted PPC.} $0.19$--$0.35$ (midpoint: $0.25$).

\noindent\textbf{Action.}
Switch to Kokkos \texttt{TeamPolicy}, SYCL \texttt{local\_accessor}, or a
vendor-tuned library (cuBLAS, rocBLAS).

\noindent\textbf{Rationale.}
R3 has the highest priority because P004 is an \emph{absolute} limitation:
no amount of problem-size scaling or granularity tuning can recover the PPC
deficit caused by a missing algorithmic primitive. R1 and R2 can be mitigated
by coarsening; R3 cannot.

\noindent\textbf{Validation.}
RAJA on E2 (DGEMM) on AMD MI300X: PPC $= 0.316$ (small) $\rightarrow 0.236$
(medium) $\rightarrow 0.189$ (large). The \emph{worsening} trend with
increasing $N$ confirms an algorithmic root cause --- overhead amortization,
which would improve PPC with $N$, is not occurring
(Section~\ref{subsec:e2}).

% --- R4 ---
\paragraph{R4: Load Imbalance Compound Check (P007).}
\textit{Priority: Second.}

\noindent\textbf{Condition.}
($\text{R1 triggered}$ \textbf{or} $T_k / \text{overhead} < 3$)
\textbf{and} $\sigma > 1.0$.

\noindent\textbf{Formula.}
\begin{equation}
    \text{PPC}_{\text{R4}} = \text{PPC}_{\text{baseline}} \times
    \bigl(1 - 0.10 \times \min(\sigma, 3.0)\bigr)
    \label{eq:r4}
\end{equation}

\noindent where $\text{PPC}_{\text{baseline}}$ is the R1 prediction if R1
triggered, or $T_k / (T_k + \text{overhead})$ otherwise.

\noindent\textbf{Action.}
Use warp-per-row CSR, segmented reductions, or native implementations for
irregular workloads at small $N$.

\noindent\textbf{Validation.}
Julia on E4 (SpMV) with power-law matrix ($\sigma = 2.5$) at small $N$ on
RTX~5060: $\text{PPC}_{\text{baseline}} = 0.56$, compound prediction
$= 0.56 \times (1 - 0.10 \times 2.5) = 0.42$; observed PPC $= 0.449$
(Section~\ref{subsec:e4}).

% --- R2 ---
\paragraph{R2: Level-Set Dispatch Budget Check (P008).}
\textit{Priority: Third.}

\noindent\textbf{Condition.}
$L \times \text{overhead}[\text{abstraction}] > 0.10 \times
T_{\text{compute}}$, where
$T_{\text{compute}} = L \times T_k$.

\noindent\textbf{Formula.}
\begin{equation}
    \text{PPC}_{\text{R2}} = \frac{T_{\text{compute}}}
    {T_{\text{compute}} + L \times \text{overhead}[\text{abstraction}]}
    \label{eq:r2}
\end{equation}

\noindent\textbf{Action.}
Prefer RAJA or Kokkos for level-set algorithms (per-launch overhead
2--5\,\textmu{}s versus Julia 7--25\,\textmu{}s or SYCL 19\,\textmu{}s).
Estimate $L \times \text{overhead}$ before committing to an abstraction.

\noindent\textbf{Validation.}
SYCL on E5 (SpTRSV) with Laplacian matrix ($L = 181$) on AMD MI300X:
$T_{\text{compute}} = 181 \times T_k$, accumulated overhead
$= 181 \times 70\,\text{\textmu{}s} = 12.67$\,ms. Predicted PPC $= 0.490$;
observed PPC $= 0.490$ --- an exact model match, the strongest quantitative
validation in this study (Section~\ref{subsec:e5}).

\noindent\textbf{Inverse case.}
When levels are \emph{wide} (many rows per level), per-launch overhead
becomes negligible relative to per-level compute time. In this regime, RAJA
and Kokkos can \emph{exceed} native PPC because their fence implementations
pipeline more efficiently than native \texttt{cudaDeviceSynchronize}:
RAJA on E5/random/large on RTX~5060 achieves PPC $= 1.299$
(Section~\ref{subsec:e5}).

% --- R1 ---
\paragraph{R1: Launch Overhead Dominance Check (P001).}
\textit{Priority: Fourth.}

\noindent\textbf{Condition.}
$T_k < \text{overhead}[\text{abstraction}]$.

\noindent\textbf{Formula.}
\begin{equation}
    \text{PPC}_{\text{R1}} = \frac{T_k}{T_k + \text{overhead}[\text{abstraction}]}
    \label{eq:r1}
\end{equation}

\noindent\textbf{Action.}
Use native implementations for kernels with $T_k < 50$\,\textmu{}s.
Coarsen task granularity (kernel fusion) to amortize overhead before adopting
an abstraction.

\noindent\textbf{Validation.}
SYCL on E1 (STREAM Triad) at small $N$ on AMD MI300X: $T_k \approx
3.8$\,\textmu{}s, overhead $\approx 19$\,\textmu{}s, predicted PPC
$= 3.8 / (3.8 + 19) = 0.167$; observed PPC $= 0.167$
(Section~\ref{subsec:e1}). Per-launch overhead values by abstraction and
backend are given in Table~\ref{tab:overhead} (Section~\ref{sec:taxonomy}).

% --- R5 ---
\paragraph{R5: Safe Abstraction Prediction (Success).}
\textit{Priority: Lowest (confirmed only when no failure rule triggers).}

\noindent\textbf{Condition.}
$T_k > 10 \times \text{overhead}[\text{abstraction}]$ \textbf{and}
R3 not triggered \textbf{and} $N$ is large.

\noindent\textbf{Predicted PPC.} $0.90$--$1.05$.

\noindent\textbf{Action.}
Abstraction is safe to adopt; near-native performance expected. RAJA and
Kokkos may exceed $1.0$ on wide-level workloads due to synchronization
pipelining advantages.

\noindent\textbf{Validation.}
RAJA on E7 (N-Body) at large $N$ on RTX~5060: $T_k / \text{overhead} = 80$,
observed PPC $= 0.981$. Kokkos on E1 (STREAM) at large $N$ on AMD MI300X:
observed PPC $= 1.016$. RAJA on E5 (SpTRSV) with random matrix at large $N$
on RTX~5060: observed PPC $= 1.299$ --- P008 inverse
(Section~\ref{subsec:e5}).

% ------------------------------------------------------------------
\subsection{Rule Evaluation Algorithm}
\label{subsec:algorithm}
% ------------------------------------------------------------------

The five rules compose into a priority cascade: the first triggered rule
determines the prediction. Algorithm~\ref{alg:predict} presents the
evaluation procedure. A reference implementation is provided in the
accompanying artifact (\texttt{analysis/decision\_framework.py}).

\begin{figure}[t]
\centering
\begin{minipage}{0.92\linewidth}
\begin{algorithmic}[1]
\Function{PredictPPC}{$\text{profile}, \text{abstraction}$}
    \State $\text{R3} \gets \Call{CheckAPIExpressivity}{\text{profile}, \text{abstraction}}$
    \If{R3.triggered}
        \State \Return R3.predicted\_ppc \Comment{P004: absolute API limitation}
    \EndIf
    \State $\text{R1} \gets \Call{CheckLaunchOverhead}{\text{profile}, \text{abstraction}}$
    \State $\text{R4} \gets \Call{CheckLoadImbalance}{\text{profile}, \text{abstraction}, \text{R1}}$
    \If{R4.triggered}
        \State \Return R4.predicted\_ppc \Comment{P007: compound $>$ individual}
    \EndIf
    \State $\text{R2} \gets \Call{CheckLevelSetBudget}{\text{profile}, \text{abstraction}}$
    \If{R2.triggered}
        \State \Return R2.predicted\_ppc \Comment{P008: accumulated dispatch}
    \EndIf
    \If{R1.triggered}
        \State \Return R1.predicted\_ppc \Comment{P001: single-kernel overhead}
    \EndIf
    \State $\text{R5} \gets \Call{CheckSafeAbstraction}{\text{profile}, \text{abstraction}, \text{R1}, \text{R3}}$
    \If{R5.triggered}
        \State \Return R5.predicted\_ppc \Comment{Success: near-native}
    \EndIf
    \State \Return $0.80$ \Comment{No rule triggered: acceptable tier}
\EndFunction
\end{algorithmic}
\end{minipage}
\captionof{algorithm}{Rule evaluation cascade for PPC prediction. Priority
order: R3 $>$ R4 $>$ R2 $>$ R1 $>$ R5. The default return of $0.80$
places unclassified configurations in the \textit{acceptable} tier.}
\label{alg:predict}
\end{figure}

\paragraph{Design decisions.}
The priority order reflects the nature of each failure mode:

\begin{itemize}
    \item \textbf{R3 first:} API limitations are absolute --- no tuning,
    scaling, or granularity change can compensate for a missing algorithmic
    primitive. Checking R3 before overhead-based rules prevents false
    positives where a large $T_k$ would mask the API deficit.

    \item \textbf{R4 before R2 and R1:} When load imbalance and launch
    overhead co-occur (P001 $\times$ P007), the compound effect produces a
    lower PPC than either rule would predict individually. R4's
    multiplicative model supersedes R1's single-factor prediction.

    \item \textbf{R5 last:} Safety is confirmed by exclusion --- only when
    no failure rule has triggered can the framework predict near-native
    performance with confidence.

    \item \textbf{Default 0.80:} Configurations where no rule triggers fall
    in the gap between clear failure and clear success. The default maps to
    the \textit{acceptable} tier boundary, signaling that the abstraction is
    likely viable but should be validated empirically.
\end{itemize}

\paragraph{Multi-abstraction recommendation.}
In practice, a developer evaluates multiple candidate abstractions. The
framework runs Algorithm~\ref{alg:predict} for each candidate and ranks
them by predicted PPC. The implementation supports four recommendation
strategies: \textsc{full\_abstraction} (best PPC $\geq 0.80$),
\textsc{abstraction\_with\_tuning} ($0.60 \leq$ best PPC $< 0.80$),
\textsc{native} (best PPC $< 0.60$), and a warning list of triggered
patterns for each candidate.

% ------------------------------------------------------------------
\subsection{Summary Lookup Table}
\label{subsec:lookup}
% ------------------------------------------------------------------

Table~\ref{tab:lookup} provides a quick-reference prediction matrix for
developers who need an immediate recommendation without running the full
rule evaluation. The table distills the five rules into predicted PPC tiers
for each combination of workload class and abstraction, derived from the
E1--E7 experimental evidence.

\begin{table}[t]
\centering
\caption{Summary lookup table: predicted PPC tier by workload class and
abstraction. Excellent: PPC $\geq 0.80$; Acceptable: $0.60$--$0.80$;
Poor: $< 0.60$. Excellent+ indicates configurations where abstraction
may exceed native (PPC $> 1.0$).}
\label{tab:lookup}
\begin{tabular}{lcccc}
\hline
\textbf{Workload Class} & \textbf{Kokkos} & \textbf{RAJA}
& \textbf{SYCL} & \textbf{Julia} \\
\hline
Memory-bound, large $N$   & Excellent  & Excellent  & Excellent  & Acceptable \\
Memory-bound, small $N$   & Acceptable & Excellent  & Poor       & Poor       \\
Compute-bound (tiled)     & Excellent  & Poor       & Excellent  & Poor       \\
Level-set, narrow levels  & Acceptable & Acceptable & Poor       & Poor       \\
Level-set, wide levels    & Excellent+ & Excellent+ & Poor       & Acceptable \\
\hline
\end{tabular}
\end{table}

Two cells warrant emphasis. First, RAJA is rated \textit{Poor} for
compute-bound tiled workloads due to R3 (P004: API Expressivity Gap) ---
RAJA's \texttt{forall} cannot express shared-memory tiling, forcing
global-memory-only access regardless of problem size. Second, SYCL is rated
\textit{Poor} for both level-set classes despite excellent single-kernel
performance (E3: 0.987, E7: 0.92) because the HSA dispatch overhead
($\approx 19$\,\textmu{}s per launch) accumulates destructively across
$L$ levels (R2/P008).

% ------------------------------------------------------------------
\subsection{Limitations and Scope}
\label{subsec:limitations}
% ------------------------------------------------------------------

The decision framework is subject to several limitations that bound its
applicability:

\paragraph{Calibration scope.}
All overhead thresholds (Table~\ref{tab:overhead}) are calibrated on the
E1--E7 kernel set across NVIDIA RTX~5060 and AMD MI300X. Generalization to
other kernels, hardware generations, or vendor platforms (e.g., Intel GPU
Max) is plausible but unvalidated. Section~\ref{sec:validation} evaluates
predictive accuracy on held-out configurations from within this experimental
matrix.

\paragraph{Post-warmup assumption.}
Overhead values represent effective device-side cost after JIT warmup.
Cold-launch overhead for Julia (150--300\,\textmu{}s on first invocation) and
SYCL kernel compilation latency are not modeled. Applications with
single-invocation kernels (e.g., initialization phases) will observe higher
overhead than predicted.

\paragraph{Unmodeled dispatch patterns.}
Julia BFS-type workloads exhibit $\approx 267$\,\textmu{}s per-level
overhead (E6 back-calculation) due to multi-kernel-per-level dispatch
(scatter + cumulative sum + gather, $\sim$10 sub-launches per level). The
current model uses 25\,\textmu{}s per level, significantly underpredicting
the overhead for graph traversal workloads. The reference implementation
flags these cases with explicit warnings.

\paragraph{Unconfirmed patterns.}
Patterns P002 (Compiler Backend Failure) and P003 (Memory Layout Mismatch)
are not covered by any decision rule. If either pattern manifests on a
target configuration, the framework will overpredict PPC for affected
abstraction--workload combinations.

\paragraph{Single-GPU scope.}
The framework targets single-GPU kernel execution. Multi-GPU execution and
distributed-memory parallelism introduce additional overhead dimensions
(inter-GPU communication, distributed load balancing, memory coherence)
that are outside the current model's scope.

\medskip
\noindent Section~\ref{sec:validation} evaluates the framework's predictive
accuracy on 39 held-out configurations, quantifying how well these five
rules generalize within the experimental matrix.
