% §2 Background
% Status: Revised 2026-03-22 — full rewrite aligned with E1–E7 results.

\section{Background}
\label{sec:background}

This section establishes the conceptual and technical vocabulary required to
interpret the methodology and results presented in this paper. We define the
performance portability metric that underpins all quantitative claims
(Section~\ref{subsec:ppc}), characterize the four abstraction layers evaluated
and their relevant properties (Section~\ref{subsec:abstraction}), describe the
target hardware architectures in terms of properties that make cross-architecture
evaluation non-trivial (Section~\ref{subsec:hardware}), and introduce
\emph{selective abstraction} as the organizing concept of this work
(Section~\ref{subsec:selective}).

% ------------------------------------------------------------------
\subsection{Performance Portability: Definition and Metrics}
\label{subsec:ppc}
% ------------------------------------------------------------------

Performance portability quantifies how consistently an application or
programming model delivers computational efficiency across a set of target
hardware platforms. Following Pennycook et al.~\cite{Pennycook2019}, we adopt
the \emph{Performance Portability Coefficient} (PPC) as the primary
quantitative metric throughout this paper. For a given application $a$,
problem $p$, and set of platforms $H$, the PPC is defined as:

\begin{equation}
    \text{PPC}(a, p, H) =
    \begin{cases}
        \dfrac{|H|}{\displaystyle\sum_{i \in H} \frac{1}{e_i(a,p)}}
        & \text{if } e_i(a,p) > 0 \; \forall \, i \in H \\[10pt]
        0 & \text{otherwise}
    \end{cases}
    \label{eq:ppc}
\end{equation}

\noindent where $e_i(a, p)$ is the \emph{performance efficiency} of
application $a$ solving problem $p$ on platform $i$, defined as the ratio
of achieved performance to the best-known performance on that platform ---
in this study, the native baseline:

\begin{equation}
    e_i(a, p) = \frac{\text{Performance}_i(a, p)}{\text{NativeBaseline}_i(p)}
    \label{eq:efficiency}
\end{equation}

The PPC is the harmonic mean of per-platform efficiencies. The harmonic mean
is deliberately chosen over the arithmetic mean: a single platform on which
an abstraction performs poorly will strongly suppress the overall PPC,
reflecting the practical reality that a portability claim is only as strong
as its weakest platform.

\paragraph{Single-platform note.}
When data covers only one platform, $|H| = 1$ and PPC$(a, p, H) = e_i(a, p)$
by definition. This is numerically correct but does not capture portability
across platforms. In this paper, SYCL data is unavailable on NVIDIA RTX~5060
for most kernels (Section~\ref{subsec:hardware}); SYCL efficiency values
are therefore single-platform and should be interpreted as AMD-only efficiency,
not cross-platform portability scores.

\paragraph{Interpretation thresholds.}
We adopt the following thresholds throughout this paper, consistent with
community practice~\cite{Pennycook2019}:

\begin{itemize}
    \item $\text{PPC} \geq 0.80$: \textbf{Excellent portability} --- abstraction
    delivers near-native performance across all platforms.
    \item $0.60 \leq \text{PPC} < 0.80$: \textbf{Acceptable portability} ---
    platform-specific tuning may be required but the abstraction is viable.
    \item $\text{PPC} < 0.60$: \textbf{Poor portability} --- the abstraction
    is failing on at least one platform; native or hybrid implementation
    should be considered.
\end{itemize}

A secondary diagnostic threshold used in this paper is the \emph{deep-profiling
trigger}: configurations where abstraction efficiency falls below
$0.85 \times \text{native}$ are flagged for root-cause analysis
(Section~\ref{sec:methodology}).

\paragraph{What PPC does not capture.}
The PPC is a necessary but not sufficient measure of abstraction value. It
does not capture: (i) absolute performance --- a framework achieving 50\% of
native on all platforms has PPC~$= 1.0$ but is uniformly slow; (ii) the
overhead \emph{structure} that explains why a given PPC is achieved; or
(iii) the dependence of efficiency on problem size, which can reverse the
ranking of abstractions across problem scales. This paper addresses all
three limitations through its four-level measurement hierarchy
(Section~\ref{sec:methodology}).

% ------------------------------------------------------------------
\subsection{Abstraction Layers: An Operational Definition}
\label{subsec:abstraction}
% ------------------------------------------------------------------

In this paper, \emph{abstraction} refers to a software layer that hides
hardware-specific implementation details behind a portable, higher-level
programming interface, allowing the same source code to target multiple GPU
architectures without modification. This is a deliberately narrow, operational
definition --- not abstraction in the general computer-science sense of
encapsulation or information hiding, but a concrete engineering choice about
where in the software stack hardware-specific decisions are made.

Every abstraction layer mediates three concerns:

\begin{enumerate}
    \item \textbf{Execution target}: the developer expresses a parallel
    pattern (e.g., \texttt{parallel\_for}, a reduction); the framework
    generates the appropriate CUDA kernel, HIP kernel, or thread pool
    depending on the backend.

    \item \textbf{Memory model}: the developer declares a data structure
    (e.g., a Kokkos \texttt{View} or a Julia \texttt{CuArray}); the
    framework handles placement and movement across host and device
    memory spaces.

    \item \textbf{Execution model}: the developer expresses algorithmic
    intent; the framework maps this to the target architecture's
    concurrency model --- NVIDIA warps, AMD wavefronts --- at compile time.
\end{enumerate}

\paragraph{The abstraction spectrum.}
The frameworks evaluated in this paper span a spectrum from zero abstraction
to full language-level abstraction:

\begin{figure}[h]
\centering
\begin{tabular}{ccccc}
\hline
\textbf{Native} & \textbf{RAJA} & \textbf{Kokkos} & \textbf{SYCL} &
\textbf{Julia} \\
CUDA / HIP & Loop-level & Data + Exec & Standard API & Language runtime \\
Max control & $\longleftarrow$ & & $\longrightarrow$ & Max portability \\
\hline
\end{tabular}
\caption{The abstraction spectrum evaluated in this paper.}
\label{fig:abstraction_spectrum}
\end{figure}

\paragraph{Kokkos~\cite{Carter2014}.}
Abstracts both execution and memory via portable \texttt{View}s and
\texttt{ExecutionSpace}s. \texttt{TeamPolicy} exposes scoped shared memory
(equivalent to CUDA \texttt{\_\_shared\_\_}), enabling shared-memory tiling
where the algorithm requires it. Widely adopted in ECP codes
($\sim$30\% per Evans et al.~\cite{Evans2022}).

\paragraph{RAJA~\cite{Hornung2014}.}
Abstracts loop-level execution policies while leaving memory management
explicit. \texttt{forall} and reduction policies are expressive for
regular loops and reductions, but RAJA provides no portable API for
scoped shared-memory allocation. Kernels that require shared-memory tiling
(e.g., DGEMM) must either fall back to vendor extensions or forgo tiling
entirely, incurring an algorithmic memory traffic penalty
(Section~\ref{sec:taxonomy}, pattern P004).

\paragraph{SYCL~\cite{Reyes2020}.}
An open Khronos standard for heterogeneous programming. \texttt{local\_accessor}
provides portable shared-memory access, making SYCL expressive for
tiling-dependent kernels. On AMD hardware, SYCL is implemented via the
AdaptiveCpp (hipSYCL) backend, which routes kernel launches through the
HSA runtime. This introduces a per-launch dispatch overhead ($\approx$19\,µs)
that is higher than native HIP or Kokkos/HIP launch costs and becomes
significant for fine-grained, level-set style workloads
(Section~\ref{sec:taxonomy}, pattern P008).

\paragraph{Julia~\cite{Bezanson2017}.}
A high-level language with just-in-time compilation via LLVM. The CUDA.jl
and AMDGPU.jl packages generate device code from high-level array operations
and custom GPU kernels written in Julia syntax. Julia arrays use
column-major storage by default, which can create coalescing advantages
(or penalties) relative to row-major native baselines depending on
access patterns. Crucially, this coalescing behavior is
architecture-specific: the presence and magnitude of the layout effect
differs between NVIDIA and AMD hardware (Section~\ref{sec:results},
pattern P005).

\paragraph{Numba (excluded).}
Numba's CUDA JIT backend does not support NVIDIA Blackwell (compute
capability~12.0) due to a PTX~9.2 / driver PTX compatibility mismatch.
Numba is excluded from all experiments. This exclusion does not affect
the generalizability of our findings, as Kokkos, RAJA, SYCL, and Julia
collectively cover the full abstraction spectrum.

% ------------------------------------------------------------------
\subsection{Target Hardware Architectures}
\label{subsec:hardware}
% ------------------------------------------------------------------

This paper evaluates abstraction effectiveness across two GPU architectures
from two vendors. Table~\ref{tab:hardware} summarizes the key characteristics
of each platform.

\begin{table}[h]
\centering
\caption{Target GPU architectures. Peak BW is manufacturer-specified;
measured BW for the RTX~5060 is 288\,GB/s (conservative, high-clock state)
due to GDDR7 dynamic clock behavior described below.}
\label{tab:hardware}
\begin{tabular}{lllllll}
\hline
\textbf{Vendor} & \textbf{GPU} & \textbf{Architecture} & \textbf{Memory}
& \textbf{Peak BW} & \textbf{Native Model} & \textbf{Role} \\
\hline
NVIDIA & RTX 5060   & Blackwell (GB107) & 8\,GB GDDR7   &  384\,GB/s & CUDA & Dev/validation \\
AMD    & MI300X     & CDNA3             & 192\,GB HBM3  & 5300\,GB/s & HIP  & Primary HPC \\
\hline
\end{tabular}
\end{table}

These two architectures differ in several ways that are directly consequential
for abstraction evaluation:

\paragraph{Thread execution width.}
NVIDIA GPUs execute threads in \emph{warps} of 32; AMD CDNA3 GPUs execute in
\emph{wavefronts} of 64. An abstraction that implicitly assumes a 32-wide
execution unit will generate suboptimal code on AMD hardware. In our E3
(3D Stencil) experiments on MI300X, enabling the wave64 execution path
recovered only 3\% of the performance gap, revealing that Kokkos'
\texttt{MDRangePolicy} code generation is the dominant bottleneck rather
than the warp-width mismatch itself (Section~\ref{sec:taxonomy}, pattern P006).

\paragraph{Memory hierarchy and shared-memory access.}
NVIDIA Blackwell provides a configurable L1/shared memory split (up to
128\,KB); AMD CDNA3 provides Local Data Share (LDS) with distinct bank
conflict behavior and different total capacity. Abstractions that do not
expose portable shared-memory allocation cannot leverage these hardware
features, forcing global-memory round-trips and degrading performance on
algorithms that require data reuse within a thread block
(Section~\ref{sec:taxonomy}, pattern P004).

\paragraph{Compiler backend maturity.}
NVIDIA's \texttt{nvcc} toolchain has over a decade of optimization history;
AMD's \texttt{hipcc} and the AdaptiveCpp SYCL backend are newer and exhibit
different code generation quality for the same abstraction-layer source.
Abstraction overhead is therefore not a property of a framework in isolation
--- it is a joint property of the framework and the compiler backend targeting
a specific architecture. This explains why PPC values for the same abstraction
can differ substantially between NVIDIA and AMD, as observed across our
experiments (Section~\ref{sec:results}).

\paragraph{RTX~5060 dynamic clock protocol.}
The RTX~5060's GDDR7 memory controller operates at a reduced clock frequency
on first access and transitions irreversibly to a high-performance state
after approximately 40 sustained high-bandwidth iterations. This behavior
is unique to consumer-grade GDDR7 and does not occur on the MI300X HBM3
stack. All RTX~5060 experiments in this paper use a 50-iteration warmup
protocol (rather than the standard 10-iteration warmup) to guarantee
that measurements are taken in the high-clock state. The reported peak
bandwidth used in roofline analysis is 288\,GB/s (the conservative
measured high-clock value), not the manufacturer-specified 384\,GB/s
theoretical peak.

% ------------------------------------------------------------------
\subsection{Selective Abstraction: Concept and Motivation}
\label{subsec:selective}
% ------------------------------------------------------------------

The standard framing of the portability problem is binary: either use an
abstraction framework (gaining portability, potentially losing performance)
or write native code (retaining performance, sacrificing portability). This
framing is empirically inadequate. Our results --- and those of prior
work~\cite{Godoy2023, Evans2022} --- demonstrate that the same abstraction
can simultaneously be excellent for one workload and catastrophic for an
adjacent one on the same hardware. Julia/CUDA.jl, for example, delivers
$1.25\times$ native DGEMM throughput on RTX~5060 via a memory layout
advantage, yet collapses to $0.35\times$ native efficiency on BFS via
dispatch amplification. The framework has not changed; the workload
structure has.

We use the term \emph{selective abstraction} to describe the principle that
optimal abstraction granularity is neither universally maximal (always use a
framework) nor universally minimal (always write native code), but is
context-dependent on three factors:

\begin{enumerate}
    \item \textbf{Workload structure}: whether the kernel is memory-bound or
    compute-bound, regular or irregular, and whether it requires
    shared-memory tiling or level-set scheduling.
    \item \textbf{Problem size}: the ratio of useful computation time to
    framework launch and dispatch overhead, which determines whether overhead
    is amortized or dominant.
    \item \textbf{Target architecture}: which backend the framework uses,
    and whether the framework's code generation matches the architecture's
    execution model.
\end{enumerate}

A key insight motivating this work is that abstraction failures are not
random. They are \emph{structural}: the same failure mode recurs across
experiments, architectures, and abstraction layers whenever the same
combination of workload structure and overhead profile is present. We call
these recurring failures \emph{abstraction failure modes} and distinguish
them from implementation bugs, driver regressions, or measurement noise.
Conversely, recurring success patterns are equally structural --- RAJA and
Kokkos consistently approach native efficiency for large, regular,
memory-bound workloads on both platforms, independent of architecture.

The central claim of this work is that these structural patterns are
measurable, classifiable, and predictable. The empirical taxonomy
(Section~\ref{sec:taxonomy}) formalizes eight such patterns (P001--P008);
the decision framework (Section~\ref{sec:framework}) translates them into
five quantified rules that replace heuristic judgment in abstraction
selection.
