% §4 Methodology
% Status: In progress

\section{Methodology}
\label{sec:methodology}

\subsection{Experimental Design}

We evaluate $5$ abstraction layers (native CUDA/HIP, Kokkos, RAJA, SYCL, Julia)
across $3$ hardware platforms and $7$ representative kernels, yielding
$315$ distinct configurations (5 abstractions $\times$ 3 platforms $\times$ 3 sizes $\times$ 7 kernels).
Each configuration is executed $30$ times, producing $9{,}450$ individual timed kernel executions.

\subsection{Measurement Protocol}

Every experiment follows a strict protocol to ensure reproducibility:

\begin{enumerate}
    \item \textbf{Hardware lock} --- GPU and memory clocks locked to maximum stable
          frequency; persistence mode enabled. Results from unlocked hardware are
          not reported.
    \item \textbf{Environment log} --- all software versions recorded before each session.
    \item \textbf{Warm-up} --- 10 executions (never recorded) to amortize JIT compilation
          (Julia), warm GPU caches, and exclude runtime initialization overhead.
    \item \textbf{Timed runs} --- 30 kernel executions, each individually logged.
    \item \textbf{Statistics} --- median $\pm$ IQR reported; outliers flagged at $>2\sigma$.
\end{enumerate}

\textbf{We never report only the mean.} GPU kernels exhibit outliers from OS jitter
and DVFS that inflate mean values. The median is the primary reported statistic.

\subsection{Performance Portability Coefficient}

We use the PPC formulation of Pennycook et al.~\cite{Pennycook2019} (Equation~\ref{eq:ppc}).
Efficiency is always relative to measured native performance, not theoretical peak.
PPC $> 0.80$ denotes excellent portability; $0.60$--$0.80$ acceptable; $< 0.60$ poor.

\subsection{Roofline Normalization}

For every configuration, we compute hardware utilization
$u = \text{BW}_\text{measured} / \text{BW}_\text{peak}$.
Native implementations must achieve $u \geq 0.80$ on the Large problem size
before abstraction experiments begin. This verifies that the environment is
correctly configured and that we are measuring abstraction overhead, not
hardware or software setup defects.

\subsection{Deep Profiling}

Any configuration where abstraction achieves $< 0.85 \times$ native performance
is automatically flagged for deep profiling. For NVIDIA, we use
Nsight Systems (\texttt{nsys}) for timeline analysis and Nsight Compute
(\texttt{ncu}) for kernel-level counters. For AMD, \texttt{rocprof} and Omniperf.
For Intel, VTune. PTX (NVIDIA) and GCN (AMD) generated code is inspected
side-by-side with the native baseline.

\subsection{Overhead Attribution}

TODO: describe the four-category attribution methodology.

\subsection{Benchmarks and Implementations}

TODO: table of 7 kernels, sources, and available abstraction implementations.
